---
title: "R tutorial China"
author: "Julia Chacón"
date: "17/2/2020"
output: html_document
---
<style>
p.caption {
  font-size: 0.8em;
}
</style>

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# TRANSPLANT EXPERIMENT GONGGA MOUNTAIN (CHINA)

This is an R tutorial for working with the Transplant Experiment data in Gongga Mountain, China. The purpose of the tutorial is to be able to download the data, read them and understand their structure prior to "play" with them.

First, is important to install all the libraries needed (1), check that indeed we have all the library needed installed (2), call these libraries (3) and download the data to the correct folders.

Here are all the steps to follow. The .R files are scripts with the codes. Open the files if you want to see how are the instructions.

## 1. Install the libraries:

```{r Libraries and data, eval=FALSE}
source("install_packages.R")
```

## 2. Check out if everything is updated (R version and libraries): 

```{r Libraries and data2, eval=FALSE}
source("setup.R")
```

## 3. Call the libraries
```{r Libraries and data3, eval=TRUE}
library("tidyverse")
library("dataDownloader")
library("DBI")
library("vegan")
library("ggvegan")
library("patchwork")
```

## 4. Download the data. 
```{r Libraries and data4, eval=TRUE}
source("instructions_to_download.R")
```

## 5. Read the data

### Cover and composition data

The script to read and process the cover data is here:

```{r Cover and composition data, eval=TRUE}
source("China/start_here.R")
```

This script is going to:
  
1. Read the data from the transplants (transplant.sqlite)
2. Load cover data and metadata (cover_thin)
3. Make a fat table (cover) with species names in columns and cover values in each plot
4. Make meta data (cover_meta) This are the meta data of the "cover" tibble

We can either source it or run it step by step to see what is doing (preferred option). If you want to run it step by step then open the file.

